Norwegian citizen science data exploration

### 2018-03-07 paper.Rmd
Fix Typo
Improved download of GBIF and population data

```{r setup, include=FALSE, echo=FALSE}
# library(plyr)

# library(rgdal)
# library(ggalt)

library(dplyr)
library(rasterVis)
library(rgbif)
library(sp)
library(raster)
library(ggplot2)

@@

coordinates_norway <- c(3500000, 3500000, 5500000, 5500000)

@@

  if (!file.exists("./data/GBIF_data.zip")) {
    download_key <- occ_download(
      paste('taxonKey =', taxon$key[1]),

      'country = NO',
      'year = 2016',
      'month >= 3',
      'month <= 5',

      'hasCoordinate = TRUE',
      type = "and"
    ) %>%
      occ_download_meta
  }

ao_data <- load_GBIF_data(download_key)
```

```{r grid, include=FALSE, echo=FALSE}}
gridsizeinkm <- 50

grid <- get_grid(gridsizeinkm, coordinates_norway)
grid_10km <- get_grid(10, coordinates_norway)

@@

```{r map_per_species, echo=FALSE, message=FALSE}

nor_map <- map_data("world", "Norway")
sp::coordinates(nor_map) <- ~long + lat ## make spp_dq_m into SpatialPointsDataFrame
sp::proj4string(nor_map) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0") ## set CRS

ggplot(as.data.frame(ao_data)) +
    annotation_map(as.data.frame(nor_map), fill="gray70", colour="gray70") +
    geom_point(aes(x=decimalLongitude, y=decimalLatitude, color=species), size=2, shape=20) +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "none") +
    coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + # limits on latitude
    xlab("Longitude") +
    ylab("Latitude") +
    facet_wrap(~ species)

@@

nor_map <- map_data("world", "Norway")
sp::coordinates(nor_map) <- ~long + lat ## make spp_dq_m into SpatialPointsDataFrame
sp::proj4string(nor_map) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0") ## set CRS
nor_map <- spTransform(nor_map, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"))

gplot(ao_data_rasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(nor_map), fill="gray40", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(nor_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + # limits on latitude
    xlab("Longitude") +
    ylab("Latitude") + 
    labs(title="Total observation density") +
    labs(subtitle=taxon['canonicalName']) +
    labs(caption = paste("Density map for all", taxon['canonicalName'], "observations in Norway, rasterized to", gridsizeinkm, "x", gridsizeinkm, "km cells."))

@@

```{r rastermap, message=FALSE, echo=FALSE}

# grid_poly <- rasterToPolygons(grid)
# 
# grid_poly_latlong <- spTransform(grid_poly, CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0"))
# 
# ao_data_rasterized <- asterize(ao_data, grid_poly_latlong, fun='count')
# 
# 
# 
# 
# 
# nor_map2 <- map_data("world", "Norway")
# sp::coordinates(nor_map2) <- ~long + lat ## make spp_dq_m into SpatialPointsDataFrame
# sp::proj4string(nor_map2) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0") ## set CRS
# 
# gplot(ao_data_rasterized2) +
#     coord_fixed() +
#     annotation_map(as.data.frame(nor_map2), fill="gray40", colour="black") +
#     geom_tile(aes(fill=log(value)), alpha=1) +
#     annotation_map(as.data.frame(nor_map2), fill=NA, colour="black") +
#     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
#     scale_fill_distiller(palette = "Spectral", na.value=NA) +
#     guides(color = guide_legend(override.aes = list(size=4))) +
#     coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + # limits on latitude
#     xlab("Longitude") +
#     ylab("Latitude")

## Human population density
```{r map_human_population, messages=FALSE, echo=FALSE}

geostat_data <- read.table("data/GeoStat_NO.csv", sep=",", header=TRUE, fill=TRUE)
geostat_data$y <- strtoi(substr(geostat_data$GRD_ID, 5, 8)) * 1000 + 500
geostat_data$x <- strtoi(substr(geostat_data$GRD_ID, 10, 13)) * 1000 + 500

sp::coordinates(geostat_data) <- ~x + y ## make spp_dq_m into SpatialPointsDataFrame
sp::proj4string(geostat_data) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs") ## set CRS

pop_unrasterized <- raster::rasterize(geostat_data, grid_10km, "TOT_P", fun='sum')

gplot(pop_unrasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(nor_map), fill="gray40", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(nor_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + # limits on latitude
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Log population") +
    ggtitle("Human population density")
```

@@

gplot(pop_rasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(nor_map), fill="gray40", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(nor_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + # limits on latitude
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Log population") +
    ggtitle("Human population density")

@@

  ggplot(num_vs_pop, aes(x=population, y=observations)) +
    geom_point()+
    geom_smooth(method=lm)
```

@@

num_vs_pop$percapita <- num_vs_pop$observations / num_vs_pop$population

sp::coordinates(num_vs_pop) <- ~x + y ## make spp_dq_m into SpatialPointsDataFrame
sp::proj4string(num_vs_pop) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs") ## set CRS

num_vs_pop_rasterized <- raster::rasterize(num_vs_pop, grid, "percapita", fun='sum')

gplot(num_vs_pop_rasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(nor_map), fill="gray40", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(nor_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + # limits on latitude
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Log observations") +
    ggtitle("Observations per capita")

# ggplot(num_vs_pop) + 
#     coord_fixed() +
#     annotation_map(as.data.frame(nor_map), fill="gray50", colour="gray50") +
#     geom_point(aes(x=x, y=y, color=log(percapita)), shape=15) +
#     scale_color_distiller(palette = "Spectral") +
#     annotation_map(as.data.frame(nor_map), fill=NA, colour="gray30") +
#     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
    #           xlim=long_dataset,  # limits on longitude
    #           ylim=lat_dataset) + # limits on latitude
   # xlab("Longitude") + ylab("Latitude")

@@

sp::coordinates(num_vs_pop) <- ~x + y ## make spp_dq_m into SpatialPointsDataFrame
sp::proj4string(num_vs_pop) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs") ## set CRS

@@

gplot(num_vs_pop_rasterized_LM) +
    coord_fixed() +
    annotation_map(as.data.frame(nor_map), fill="gray40", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(nor_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) + # limits on latitude
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Log observations") +
    ggtitle("Observations relative to regression with human population")

# ggplot(num_vs_pop) + 
#     annotation_map(world_map, fill="gray50", colour="gray50") +
#     geom_point(aes(x=long_rounded, y=lat_rounded, color=log(relativeToLM)), shape=15, size=cellsize) +
#     scale_color_distiller(palette = "Spectral") +
#     annotation_map(world_map, fill=NA, colour="gray30") +
#     theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
#     coord_map(projection = "gnomonic", orientation=c(60, 16, -5),
#               xlim=long_dataset,  # limits on longitude
#               ylim=lat_dataset) + # limits on latitude
#    xlab("Longitude") + ylab("Latitude")

>>>>

```{r setup, include=FALSE, echo=FALSE}
  library(dplyr)
  library(rasterVis)
  library(rgbif)
  library(sp)
  library(raster)
  library(ggplot2)

@@

  country_code <- "NO"
  country_coordinates <- c(3500000, 3500000, 5500000, 5500000)
  country_map <- map_data("world", country_code)

  ## Season onsets for Værnes airport according to Norwegian Meteorological Institute
  spring <- as.Date("3-15",  format = "%m-%d") ## From Mar 15 
  summer <- as.Date("5-22",  format = "%m-%d")  ## From May 22
  fall <- as.Date("9-11",  format = "%m-%d")  ## From Sep 11
  winter <- as.Date("11-18",  format = "%m-%d")  ## From Nov 18

  sp::coordinates(country_map) <- ~long + lat
  sp::proj4string(country_map) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0")
```

@@

  gbif_filter <- list(
      paste('taxonKey =', taxon$key[1]),
      paste('country =', country_code),
      'datasetKey = b124e1e0-4755-430f-9eab-894f25a9b59c',
      'year = 2016',
      'hasCoordinate = TRUE',
      'hasGeospatialIssue = FALSE',
      type = "and")

  ao_data <- load_GBIF_data(gbif_filter)
  geostat_data <- load_GeoStat_data(country_code)

```{r grid, include=FALSE, echo=FALSE}
  gridsizeinkm <- 50

  grid <- get_grid(gridsizeinkm, country_coordinates)
  grid_10km <- get_grid(10, country_coordinates)

@@

```{r map_per_species, echo=FALSE, message=FALSE}
  ggplot(as.data.frame(ao_data)) +
    annotation_map(as.data.frame(country_map), fill="gray70", colour="gray70") +
    geom_point(aes(x=decimalLongitude, y=decimalLatitude, color=species), size=.5, shape=20) +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "none") +
    coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    facet_wrap(~ species)

@@

  country_map <- spTransform(country_map, CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"))

  gplot(ao_data_rasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="gray40", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") + 
    labs(title="Total observation density") +
    labs(subtitle=taxon['canonicalName']) +
    labs(caption = paste("Density map for all", taxon['canonicalName'], "observations in Norway, rasterized to", gridsizeinkm, "x", gridsizeinkm, "km cells."))

@@

## Human population density
```{r map_human_population, messages=FALSE, echo=FALSE}
  pop_unrasterized <- raster::rasterize(geostat_data, grid_10km, "TOT_P", fun='sum')

  gplot(pop_unrasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="gray40", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Log population") +
    ggtitle("Human population density")
```

@@

gplot(pop_rasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="gray40", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Log population") +
    ggtitle("Human population density")

@@

  # num_vs_pop <- subset(num_vs_pop, population<100000 & observations < 1000)

  sp::coordinates(num_vs_pop) <- ~x + y
  sp::proj4string(num_vs_pop) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")

  ggplot(as.data.frame(num_vs_pop), aes(x=population, y=observations)) +
    geom_point()+
    geom_smooth(method=lm)
```

@@

num_vs_pop$percapita <- num_vs_pop$observations / num_vs_pop$population
num_vs_pop_rasterized <- raster::rasterize(num_vs_pop, grid, "percapita", fun='sum')

gplot(num_vs_pop_rasterized) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="gray40", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Log observations") +
    ggtitle("Observations per capita")
```

@@

sp::coordinates(num_vs_pop) <- ~x + y
sp::proj4string(num_vs_pop) <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs")

@@

gplot(num_vs_pop_rasterized_LM) +
    coord_fixed() +
    annotation_map(as.data.frame(country_map), fill="gray40", colour="black") +
    geom_tile(aes(fill=log(value)), alpha=1) +
    annotation_map(as.data.frame(country_map), fill=NA, colour="black") +
    theme(panel.background = element_rect(fill="aliceblue"), legend.position = "right") +
    scale_fill_distiller(palette = "Spectral", na.value=NA) +
    guides(color = guide_legend(override.aes = list(size=4))) +
    # coord_map(projection = "gnomonic", orientation=c(60, 16, -5)) +
    xlab("Longitude") +
    ylab("Latitude") +
    labs(fill = "Log observations") +
    ggtitle("Observations relative to regression with human population")
```
